---
title: "softwareRisk: Computation of Node and Path-Level Risk Scores in Scientific Models"
output: rmarkdown::pdf_document
bibliography: references_pdf.bib
vignette: >
  %\VignetteIndexEntry{softwareRisk: Computation of Node and Path-Level Risk Scores in Scientific Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2
)
```

The `R` package `softwareRisk` leverages the network-like architecture of scientific models 
together with software quality metrics to identify risky paths, which are defined by the complexity
of its functions and the extent to which errors can cascade along and beyond their 
execution order. It operates on `tbl_graph` objects representing 
call dependencies between functions (callers and callees). By leveraging the `sensobol` 
package [@puy2022a], `softwareRisk` also supports variance-based uncertainty and sensitivity analyses
to evaluate how the identification of risky function-call paths varies under alternative assumptions 
about the relative importance of function complexity, coupling and structural position within 
the software.

# Workflow

We first load the packages needed for the analysis.

```{r setup, message=FALSE}
library(softwareRisk)
library(tidygraph)
```

## Prepare the required datasets

`softwareRisk` draws on the representation of the model's source code as a directed call 
graph $G = (V,E)$ in which each node $v_i \in V$ is a function or subroutine and each 
directed edge $e_{ij} = (v_i, v_j) \in E$ is a function call. It also assumes that 
each function will have a cyclomatic complexity value. Therefore the analyst should have two 
different datasets before starting the analysis:

1. A spreadsheet listing the set set of directed edges as an edge list, with one row per
function call. The first column may contian the caller function (source node, $v_i$, "from") and the
second column the calle function (target node, $v_j$, "to"): 

2. A spreadsheet listing the cyclomatic_complexity values for each function in the model.

Let us create these datasets to illustrate their format:

```{r}

# Dataset 1: calls (edge list) -------------------------------------------------

calls_df <- data.frame(
  from = c("clean_data", "compute_risk", "compute_risk", "calc_scores", "plot_results"),
  to = c("load_data", "clean_data", "calc_scores", "mean", "compute_risk")
)

calls_df

# Dataset 2: cyclomatic complexity (node attributes) ---------------------------

cyclo_df <- data.frame(
  name  = c("clean_data", "load_data", "compute_risk", "calc_scores", "mean", "plot_results"),
  cyclo = c(6, 3, 12, 5, 2, 4)
)

cyclo_df
```

The analyst can then merge them into a `tbl_graph`:

```{r merge}

# Merge into a tbl_graph -------------------------------------------------------

graph <- tbl_graph(nodes = cyclo_df, edges = calls_df, directed = TRUE)

graph
```

Once this is done, the data is prepared for `softwareRisk`.

## Analysis

Here we illustrate the functions of `softwareRisk` by using the build-in data
`synthetic_graph`. It consists of five entry nodes, 35 middle nodes and
15 sink nodes. Each entry node calls between two and five middle nodes and each
middle node calls one to three sink nodes, thus simulating realistic code architecture. 
The synthetic example reproduces the characteristic right-tailed distribution of cyclomatic complexity 
found in real software systems, with many low-complexity functions and few highly complex ones [@landman2016].

```{r data_loading}

# Load the data ----------------------------------------------------------------

data("synthetic_graph")

# Print it ---------------------------------------------------------------------

synthetic_graph
```

The next step is to compute the in-degree and betweenness centrality of each node, calculate its 
risk score and identify all simple paths through the directed function-call graph. All this
is done with the ``all_paths_fun`` function. The in-degree and betweenness of the nodes
are calculated internally by functions in the `igraph` package. 

### Definition of software risk

Risk is computed using a weighted power-mean aggregation of normalized cyclomatic complexity, in-degree and betweenness. The power parameter $p$ controls how attributes combine: values of $p < 1$ emphasize functions where several risk factors co-occur, while values of $p > 1$ increasingly focus on the largest individual contributor. When $p = 1$, the formula reduces to a simple weighted sum (additive risk).

\begin{equation}
r^{(p)}_{(v_i)} =
\left(
\alpha\, \tilde{C}_{(v_i)}^{p}
+ \beta\, \tilde{d}_{(v_i)}^{\mathrm{in}\,p}
+ \gamma\, \tilde{b}_{(v_i)}^{p}
\right)^{1/p},
\qquad p \in [-1,2].
\label{eq:composite_risk_score_power}
\end{equation}

where the tilde $\tilde{}$ refers to normalization, $C$ denotes cyclomatic complexity, $d^{\text{in}}$ refers
to in-degree and $b$ denotes betweenness. The weights $\alpha$, $\beta$ and $\gamma$ reflect the relative importance of complexity, coupling and network position and can be defined by the analyst, with the
constraint that $\alpha+\beta+\gamma =1$. High $r$ values indicate functions that are complex and/or highly interconnected and hence potential points of structural vulnerability. Table 1 summarises the interpretation of each parameter.

| Parameter | Attribute | Risk type | Definition of risk | Intuition |
|---|---|---|---|---|
| $\alpha$ | Cyclomatic complexity | Complexity-driven | Risk arises from internal complexity | Hard-to-read functions are more error-prone |
| $\beta$ | In-degree | Dependency-driven | Risk arises from widespread reuse | Highly reused functions can propagate failures |
| $\gamma$ | Betweenness | Structural-driven | Risk arises from structural centrality | Bridge functions can disrupt large parts of the model |
| $p<1$ | - | Compensatory | Attributes offset one another | Weaknesses can compensate |
| $p=1$ | - | Additive | Risk is a weighted sum | Effects add linearly |
| $p>1$ | - | Reinforcing | Attributes reinforce each other | Vulnerabilities amplify |
| $p\gg 1$ | - | Bottleneck | Risk set by the largest attribute | Worst case dominates |

Table: Interpretations of the parameters forming the risk score. For the weights, $(\alpha, \beta, \gamma) \in [0, 1]$, and $\alpha + \beta + \gamma = 1$.

## Path-level risk scores

The risk scores computed at the function level are then aggregated at the path level as

\begin{equation}
P_k = 1 - \prod_{i=1}^{n_k} (1 - r_{k(v_i)})\,,
\label{eq:independent_events}
\end{equation}

where $r_{k(v_i)}$ is the risk of the $i$-th function in path $k$ and $n_k$ is the number of functions in that path. $P_k$ is at least as large as the maximum individual function risk and monotonically increases as more functions on the path become risky, approaching 1 when several functions have high risk scores. High $P_k$ scores thus identify not only vulnerable paths, but also paths whose potential failure can have a larger cascading effect into other parts of the system through their shared high-centrality functions.

In this example, we set $\alpha=0.6$, $\beta=0.3$ and $\gamma = 0.1$, thus prioritizing defect-proneness and the likelihood of unexpected behaviours and relegating propagation potential as secondary. By default, $p=1$ in the ``all_paths_fun`` (additive risk).

```{r all_paths}

# Run the function -------------------------------------------------------------

output <- all_paths_fun(graph = synthetic_graph,
                        alpha = 0.6,
                        beta  = 0.3,
                        gamma = 0.1,
                        complexity_col = "cyclo",
                        weight_tol = 1e-8)

# Print the output -------------------------------------------------------------

output
```

The output of `all_paths_fun` is a list with two slots, `nodes` and `paths`. 

* `$nodes`: it contains the name of the nodes and their relevant metrics (cyclomatic complexity, in-degree, out-degree, betweenness and risk score).

* `$paths`: it describes each simple path in the call graph and includes the ordered sequence of nodes forming the path, the number of hops (i.e., the number of edges traversed along the path), the path-level risk score and the vector of cyclomatic complexity values for the nodes along the path (`path_cc` column). In addition, two distributional metrics are reported: the Gini coefficient of node-level risk along the path (`gini_node_risk` column), which quantifies the inequality of risk contributions across nodes within a path, and the risk_slope, which captures the direction and magnitude of change in node-level risk from the beginning to the end of the path.

### Plotting

`softwareRisk` provides functions to inspect the results of the analysis. The function `path_fix_heatmap` allows the analyst to chose the top $n$ nodes and $n$ paths in terms of their risk score and 
observe how much the risk score of the riskiest paths would decrease if the selected high-risk nodes were made perfectly reliable. This analysis identifies nodes that act as chokepoints for risk propagation, highlights paths dominated by single high-risk functions and reveals which refactoring actions would yield the greatest reductions in path-level risk.

```{r plot_heatmap, fig.height=2.5, fig.width=3.5}

path_fix_heatmap(all_paths_out = output, n_nodes = 20, k_paths = 20)
```

`softwareRisk` also allows to plot the call graph with the top risky paths highlighted. This is done with the function `plot_top_paths_fun`. The top ten most risky paths are highlighted in colour. The thickness of the edge shows how frequently an edge participates in the top 10 most risky paths. The color of the edge (from orange to red) indicates the mean risk of paths including that edge.

```{r plot_paths, fig.height=2.5, fig.width=3.5}

plot_output <- plot_top_paths_fun(graph = synthetic_graph,
                                  all_paths_out = output,
                                  model.name = "ToyModel",
                                  language = "Fortran",
                                  top_n = 10,
                                  alpha_non_top = 0.05)
```

The color of the nodes maps onto the cyclomatic complexity categories defined by @watson1996 (0-10 low risk; 10-20 moderate complexity, 20-50 complex, high risk; $>50$ very complex, untestable).

The `alpha_non_top` argument controls the transparency of the paths that are not identified as top. For small or sparse models, it can be set to `alpha_non_top = 1` to better visualize the full call graph:

```{r plot_paths2, fig.height=2.5, fig.width=3.5}

plot_output <- plot_top_paths_fun(graph = synthetic_graph,
                                  all_paths_out = output,
                                  model.name = "ToyModel",
                                  language = "Fortran",
                                  top_n = 10,
                                  alpha_non_top = 1)
```

### Uncertainty and sensitivity analysis

`softwareRisk` also enables the analyst to perform uncertainty and sensitivity analyses of risk and path score calculations by leveraging the sensobol package [@puy2022a]. By systematically varying the weights $(\alpha, \beta, \gamma)$ and the aggregation parameter $p$, the package allows users to evaluate how sensitive node- and path-level risk scores are to different risk conceptualizations. This approach makes it possible to assess the robustness of the identification of high-risk paths under alternative definitions of risk.

Uncertainty and sensitivity analyses are implemented through the function `uncertainty_fun`. The user needs to define the order of the effects explored (`first`, `second` or `third`).

Internally, `uncertainty_fun` builds a Sobol' quasi-random design over four independent $U(0,1)$ draws. Three of them (`a_raw`, `b_raw`, `c_raw`) are normalised to sum to one, yielding the weights $\alpha$, $\beta$ and $\gamma$; the fourth (`p_raw`) is mapped linearly to $p \in [-1, 2]$. Independent uniform inputs are required by the quasi-random sequence, hence the need for the raw draws; the sensitivity indices, however, are attributed to the transformed parameters and labelled `alpha`, `beta`, `gamma` and `p` in the output, so the results are directly interpretable in terms of the model parameters.

```{r uncertainty}

# Run uncertainty analysis -----------------------------------------------------

uncertainty_analysis <- uncertainty_fun(all_paths_out = output,
                                        N = 2^10,
                                        order = "first")

# Print the top five rows ------------------------------------------------------

lapply(uncertainty_analysis, function(x) head(x, 5))
```

The output is a list with two slots:

* `$nodes`: a `name` column with the name of the node, an `uncertainty_analysis` column with a vector of $N$ node-level risk scores after randomizing $\alpha$, $\beta$, $\gamma$ and $p$, and a `sensitivity_analysis` column with the results of the sensitivity analysis. Each element of `sensitivity_analysis` is a `sensobol::sobol_indices()` object whose `$results` data frame reports first-order ($S_i$) and/or total-order ($T_i$) indices for the four parameters, labelled `alpha`, `beta`, `gamma` and `p`. See the `sensobol` package for further details [@puy2022a].

* `$paths`: a `path_id` column with the ID number of the path, a `path_str` column informing on the sequence of functions calls for that path, a `hops` column informing on the number of edges traversed along the path and three columns informing on the results of the uncertainty analysis (UA): 
  - `uncertainty_analysis`: vector of path-level risk scores after the UA.
  - `gini_index`: vector of gini_index values after the UA.
  - `risk_trend`: vector of risk_slope values after the UA.

The analyst can also plot the top $n$ risky paths and their uncertainty with the function `path_uncertainty_plot`. The error bars encompass the minimum, mean and maximum $P_k$ value for that path after the uncertainty analysis.

```{r plot_uncert, fig.height=2.5, fig.width=3.5}

path_uncertainty_plot(ua_sa_out = uncertainty_analysis, n_paths = 20)
```

### Extracting and plotting Sobol' indices

The `sensitivity_analysis` list-column in `$nodes` stores one `sensobol::sobol_indices()` object per node. The Sobol' indices for a given node are accessible via the `$results` slot of that object, which is a data frame with three columns: `original` (the index value), `sensitivity` (`"Si"` for first-order, `"Ti"` for total-order), and `parameters` (`"alpha"`, `"beta"`, `"gamma"`, `"p"`).

```{r sa_single_node}

# Sobol' indices for the first node
si_node1 <- uncertainty_analysis$nodes$sensitivity_analysis[[1]]$results
si_node1
```

To compare parameter importance across all nodes, combine the per-node results into a single data frame:

```{r sa_combine}

sa_all <- do.call(rbind, Map(
  function(sa, nm) data.frame(sa$results, name = nm, stringsAsFactors = FALSE),
  uncertainty_analysis$nodes$sensitivity_analysis,
  uncertainty_analysis$nodes$name
))

head(sa_all)
```

The resulting data frame can be used to visualize how much of the variance in node risk scores is explained by each parameter, and whether the dominant driver is consistent across nodes or varies. The plot below shows boxplots of $S_i$ and $T_i$ for each parameter across all nodes. A large gap between $S_i$ and $T_i$ for a parameter indicates important higher-order interactions with the other parameters.

```{r sa_plot, fig.height=2.5, fig.width=4}

library(ggplot2)

ggplot(sa_all, aes(x = parameters, y = original, fill = sensitivity)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(
    values = c(Si = "#F8766D", Ti = "#00BFC4"),
    labels = c(expression(S[i]), expression(T[i]))
  ) +
  labs(x = "Parameter", y = "Sobol' index", fill = "Index") +
  theme_bw()
```

## References

